\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-11-12 archive.tex
symmetry breaking section

Together with its weight vector, a DSL defines a distribution over
programs, $\probability[p|\mathcal{D},\theta]$.  We define this
distribution by specifying a procedure for drawing samples from
$\probability[p|\mathcal{D},\theta]$ (Algorithm~\ref{sampleProgram}).
Care must be taken to ensure that variable scoping rules are obeyed
and that programs are well-typed.  We ensure well-typed programs by
performing Hindley-Milner type inference~\cite{pierce} during
sampling, and assume that each task is annotated with the type of the
program that will solve it.
 \begin{algorithm}
   \caption{Generative model over programs}
   \label{sampleProgram}
   \begin{algorithmic}[1]
     \State \textbf{function} sample$(\mathcal{D}, \theta, \tau)$:
     \State {\bfseries Input:} DSL $(\mathcal{D},\theta)$, type $\tau$
     \State \textbf{Output:} a program whose type unifies with $\tau$
     \State  \textbf{return }sample$'(\mathcal{D}, \theta, \varnothing, \tau)$

@@

Consider an agent tasked with writing programs built from addition, multiplication, and the constants zero and one.
A bigram parameterization of $Q$ allows
it to represent the fact that it should never add zero ($Q_{\code{+},\code{0},0} = Q_{\code{+},\code{0},1} = 0$)
or never multiply by  1 ($Q_{\code{*},\code{1},0} = Q_{\code{*},\code{1},1} = 0$), or
that addition should always associate to the right
($Q_{\code{+},\code{+},0} = 0$).
The $\mathcal{L}^{\text{MAP}}$ training objective encourages
learning these canonical forms.
Consider two recognition models, $Q_1$ and $Q_2$,
and two programs in frontier $\mathcal{F}_x$,
$p_1=\code{(+ (+ 1 1) 1)}$ and $p_2=\code{(+ 1 (+ 1 1))}$,
where
\begin{align*}
  Q_1(p_1|x) = \frac{\epsilon}{2}&\qquad Q_1(p_2|x) = \frac{\epsilon}{2}\\
  Q_2(p_1|x) = 0&\qquad  Q_2(p_2|x) = \epsilon
\end{align*}
i.e., $Q_2$ breaks a symmetry by forcing right associative addition,
but $Q_1$ does not, instead splitting its probability mass equally between $p_1$ and $p_2$.
Now because $\probability[p_1|\mathcal{D},\theta] = \probability[p_2|\mathcal{D},\theta]$
(Algorithm~\ref{sampleProgram}),
we have
\begin{align*}
  \mathcal{L}^{\text{posterior}}_{\text{real}}(Q_1)& = \frac{\probability[p_1|\mathcal{D},\theta]\log \frac{\epsilon}{2} + \probability[p_2|\mathcal{D},\theta]\log \frac{\epsilon}{2}}{\probability[p_1|\mathcal{D},\theta] + \probability[p_2|\mathcal{D},\theta]} = \log \frac{\epsilon}{2}\\
  \mathcal{L}^{\text{posterior}}_{\text{real}}(Q_2)& = \frac{\probability[p_1|\mathcal{D},\theta]\log 0 + \probability[p_2|\mathcal{D},\theta]\log \epsilon}{\probability[p_1|\mathcal{D},\theta] + \probability[p_2|\mathcal{D},\theta]} = +\infty\\
  \mathcal{L}^{\text{MAP}}_{\text{real}}(Q_1)& = \log Q_1(p_1)           = \log Q_1(p_2)   = \log \frac{\epsilon}{2}\\
  \mathcal{L}^{\text{MAP}}_{\text{real}}(Q_2)& = \log Q_2(p_2) \phantom{ = \log Q_1(p_2) } = \log \epsilon\\
\end{align*}
So the $\mathcal{L}^{\text{MAP}}$ prefers $Q_2$ (the symmetry breaking recognition model),
while $\mathcal{L}^{\text{posterior}}$ reverses this preference.
We can confirm this experimentally by training
four recognition models:
a

@@

\begin{figure*}
  \centering\begin{tikzpicture}[every node/.style={inner sep=1,outer sep=0}]
    \node(p1)[draw] at (-1,0) {
      \begin{tabular}{l}
      \texttt{(Y ($\lambda$ (r l) (if (nil? l) nil}\\
      \texttt{ (cons (+ (car l) (car l))}\\
      \phantom{\texttt{(cons }}\texttt{ (r (cdr l))))))}
    \end{tabular}
    };
    \node(r1)[draw,inner sep=0,outer sep=0] at ([yshift=-3cm]p1.south) {
      \begin{tabular}{l}
  \texttt{(}\orange{\texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if (nil? l)}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{nil}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{(cons (f (car l))}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{ (r (cdr l)))))))}}\\
\texttt{ ($\lambda$ (z) (+ z z)))}
      \end{tabular}
    };

@@

    };
    \node(r2)[draw] at ([yshift=-3cm]p2.south) {
\begin{tabular}{l}
  \texttt{(}\orange{\texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if (nil? l)}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{nil}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{(cons (f (car l))}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{ (r (cdr l)))))))}}\\
\texttt{ ($\lambda$ (z) (- z 1)))}
      \end{tabular}

@@

    \node[draw](m) at (3,-7) {
\begin{tabular}{l}
          \fbox{\textsc{map}} = \orange{\texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if (nil? l) nil}}\\
  \phantom{\texttt{\emph{map}} = \texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if }}\orange{\texttt{(cons (f (car l))}}\\
  \phantom{\texttt{\emph{map}} = \texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if }}\orange{\texttt{(r (cdr l))))))}}\\
  \code{(}\fbox{\textsc{map}}\code{ ($\lambda$ (z) (+ z z)))}\quad(program rewritten w/ new \fbox{\textsc{map}} primitive)\\
  \code{(}\fbox{\textsc{map}}\code{ ($\lambda$ (z) (- z 1)))}\quad(program rewritten w/ new \fbox{\textsc{map}} primitive)\\
      \end{tabular}      
    };
    \draw [->](r1.south)--(m.north);
    \draw [->](r2.south)--(m.north);
    \node[fill=white] at ([yshift=0.5cm]m.north) {\textbf{Compress (MDL/Bayes objective)}};
    %% \visible<2-3>{
    %%   \node[fill=orange!50!white,align=center] at (r1) {\Huge $10^{14}$\\\Huge  refactorings };
    %%   \node[fill=orange!50!white,align=center] at (r2) {\Huge $10^{13}$\\\Huge  refactorings };
    %% }
    %% \visible<3>{
    %%   \node[fill=orange!50!white,align=center] at ([yshift=7cm]m.north) {\Huge version space: set of programs\\
    %%   \huge Lau 2003; Gulwani 2012};
    %%   }
    %% \visible<4>{
    %%   \node[fill=orange!50!white,align=center] at (r1) {\Huge $\leq 10^6$\\\Huge version spaces };
    %%   \node[fill=orange!50!white,align=center] at (r2) {\Huge $\leq 10^6$\\\Huge version spaces };
    %% }

@@

>>>>

Together with its weight vector, a DSL defines a distribution over
programs, $\probability[p|\mathcal{D},\theta]$.  We define this
distribution by specifying a procedure for drawing samples from
$\probability[p|\mathcal{D},\theta]$ (Algorithm~\ref{sampleProgram}).
Care must be taken to ensure that variable scoping rules are obeyed
and that programs are well-typed.  We ensure well-typed programs by
performing Hindley-Milner type inference~\cite{pierce} during
sampling, and assume that each task is annotated with the type of the
program that will solve it.
Appendix~\ref{enumerationAppendix}
explains how we enumerate,
rather than sample,
programs generated by Algorithm~\ref{sampleProgram}.
 \begin{algorithm}
   \caption{Generative model over programs}
   \label{sampleProgram}
   \begin{algorithmic}[1]
     \State \textbf{function} sample$(\mathcal{D}, \theta, \tau)$:
     \State {\bfseries Input:} DSL $(\mathcal{D},\theta)$, type $\tau$
     \State \textbf{Output:} a program whose type unifies with $\tau$
     \State  \textbf{return }sample$'(\mathcal{D}, \theta, \varnothing, \tau)$

@@

Consider an agent tasked with writing programs built from addition and the constants zero and one.
A bigram parameterization of $Q$ allows
it to represent the fact that it should never add zero ($Q_{\code{+},\code{0},0} = Q_{\code{+},\code{0},1} = 0$)
%or never multiply by  1 ($Q_{\code{*},\code{1},0} = Q_{\code{*},\code{1},1} = 0$), or
or that addition should always associate to the right
($Q_{\code{+},\code{+},0} = 0$).
The $\mathcal{L}^{\text{MAP}}$ training objective encourages
learning these canonical forms.
Consider two recognition models, $Q_1$ and $Q_2$,
and two programs in frontier $\mathcal{F}_x$,
$p_1=\code{(+ (+ 1 1) 1)}$ and $p_2=\code{(+ 1 (+ 1 1))}$,
where
\begin{align*}
  Q_1(p_1|x) = \frac{\epsilon}{2}&\qquad Q_1(p_2|x) = \frac{\epsilon}{2}\\
  Q_2(p_1|x) = 0&\qquad  Q_2(p_2|x) = \epsilon
\end{align*}
i.e., $Q_2$ breaks a symmetry by forcing right associative addition,
but $Q_1$ does not, instead splitting its probability mass equally between $p_1$ and $p_2$.
Now because $\probability[p_1|\mathcal{D},\theta] = \probability[p_2|\mathcal{D},\theta]$
(Algorithm~\ref{sampleProgram}),
we have
\begin{align*}
  \mathcal{L}^{\text{posterior}}_{\text{real}}(Q_1)& = \frac{\probability[p_1|\mathcal{D},\theta]\log \frac{\epsilon}{2} + \probability[p_2|\mathcal{D},\theta]\log \frac{\epsilon}{2}}{\probability[p_1|\mathcal{D},\theta] + \probability[p_2|\mathcal{D},\theta]} = \log \frac{\epsilon}{2}\\
  \mathcal{L}^{\text{posterior}}_{\text{real}}(Q_2)& = \frac{\probability[p_1|\mathcal{D},\theta]\log 0 + \probability[p_2|\mathcal{D},\theta]\log \epsilon}{\probability[p_1|\mathcal{D},\theta] + \probability[p_2|\mathcal{D},\theta]} = +\infty\\
  \mathcal{L}^{\text{MAP}}_{\text{real}}(Q_1)& = \log Q_1(p_1)           = \log Q_1(p_2)   = \log \frac{\epsilon}{2}\\
  \mathcal{L}^{\text{MAP}}_{\text{real}}(Q_2)& = \log Q_2(p_2) \phantom{ = \log Q_1(p_2) } = \log \epsilon\\
\end{align*}
So $\mathcal{L}^{\text{MAP}}$ prefers $Q_2$ (the symmetry breaking
recognition model), while $\mathcal{L}^{\text{posterior}}$ reverses
this preference.  We experimentally confirm this symmetry-breaking
behavior by training recognition models that minimize either
$\mathcal{L}^{\text{MAP}}$/$\mathcal{L}^\text{posterior}$ and which
use either a bigram parameterization/unigram\footnote{In the unigram variant $Q$ predicts a $|\mathcal{D}| + 1$-dimensional vector: $Q(p|x) = \probability[p|\mathcal{D},\theta_i = Q_i(x)]$,
  and was used in our prior work~\cite{ecc}} parameterization.
Figure~\ref{symmetry} shows the result of training $Q$ in these four regimes for a DSL containing \code{+}, \code{0}, and \code{1}
and then sampling programs.
On this particular run,
the combination of
bigrams and $\mathcal{L}^{\text{MAP}}$ learns to
avoid adding zero and associate addition to the right ---
different random initializations
lead to either right or left association.

To be clear,
our recognition model does not learn to break \emph{every} possible symmetry in every possible DSL.
But in practice we found that a bigrams combined with $\mathcal{L}^{\text{MAP}}$
works well,
and we use with this combination throughout the rest of the paper.
\begin{figure}
\centering  \begin{tabular}{ccc}
     \toprule
&     Unigram&Bigram \\\midrule
     $\mathcal{L}^{\text{posterior}}$&
     \begin{tabular}{l}
       \emph{Three samples:}\\
       \code{(+ 1 0)}\\
\code{(+ (+ 0 0) (+ 1 0))}\\
\code{(+ 1 1)}
\\
63.0\% right-associative; 37.4\% \code{+0}'s
       \end{tabular}
     &
     \begin{tabular}{l}
       \emph{Three samples:}\\
       \code{0}\\
\code{(+ (+ (+ 0 0) (+ 0 1)) 1)}\\
\code{1}\\
55.8\% right-associative; 31.9\% \code{+0}'s
       \end{tabular}
     \\\\
     $\mathcal{L}^{\text{MAP}}$&
     \begin{tabular}{l}
       \emph{Three samples:}\\
       \code{1}\\
       \code{(+ 1 (+ 1 (+ (+ 1 (+ 1 1)) 1)))}\\
       \code{(+ (+ 1 1) 1)}
\\48.6\% right-associative;
       0.5\% \code{+0}'s       
       \end{tabular}
     &
     \begin{tabular}{l}
       \emph{Three Samples:}\\
\code{(+ 1 (+ 1 (+ 1 (+ 1 (+ 1 1)))))}\\
\code{0}\\
\code{(+ 1 (+ 1 (+ 1 1)))}
       \\
       \textbf{97.9\% right-associative};
       2.5\% \code{+0}'s       
       \end{tabular}
\\    \bottomrule 
  \end{tabular}
  \caption{Agent learns to break symmetries in program space only when using both bigram parameterization and $\mathcal{L}^{\text{MAP}}$ objective, associating addition to the right and avoiding adding zero. \% right-associative calculated by drawing 500 samples from $Q$. $\mathcal{L}^{\text{MAP}}$/Unigram agent incorrectly learns to never generate programs with \code{0}'s, while $\mathcal{L}^{\text{MAP}}$/Bigram agent correctly learns that \code{0} should only be disallowed as an argument of addition. Tasked with building programs from \code{+}, \code{1}, and \code{0}. }\label{symmetry}
  \end{figure}

@@

\begin{figure*}
  \centering\begin{tikzpicture}[every node/.style={inner sep=1,outer sep=0}]
    \node(p1)[draw] at (-1,0) {
      \begin{tabular}{l}
      \texttt{(Y ($\lambda$ (r l) (if (nil? l) nil}\\
      \texttt{ (cons (+ (car l) (car l))}\\
      \phantom{\texttt{(cons }}\texttt{ (r (cdr l))))))}
    \end{tabular}
    };
    \node(r1)[draw,inner sep=0,outer sep=0] at ([yshift=-2.5cm]p1.south) {
      \begin{tabular}{l}
  \texttt{(}\orange{\texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if (nil? l)}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{nil}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{(cons (f (car l))}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{ (r (cdr l)))))))}}\\
\texttt{ ($\lambda$ (z) (+ z z)))}
      \end{tabular}
    };

@@

    };
    \node(r2)[draw] at ([yshift=-2.5cm]p2.south) {
\begin{tabular}{l}
  \texttt{(}\orange{\texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if (nil? l)}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{nil}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{(cons (f (car l))}}\\
\phantom{(($\lambda$ (f) (Y ($\lambda$ (r l)}\orange{\texttt{ (r (cdr l)))))))}}\\
\texttt{ ($\lambda$ (z) (- z 1)))}
      \end{tabular}

@@

    \node[draw](m) at (3,-6.5) {
\begin{tabular}{l}
          \fbox{\textsc{map}} = \orange{\texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if (nil? l) nil}}\\
  \phantom{\texttt{\emph{map}} = \texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if }}\orange{\texttt{(cons (f (car l))}}\\
  \phantom{\texttt{\emph{map}} = \texttt{($\lambda$ (f) (Y ($\lambda$ (r l) (if }}\orange{\texttt{(r (cdr l))))))}}\\
  \code{(}\fbox{\textsc{map}}\code{ ($\lambda$ (z) (+ z z)))}\quad(program rewritten w/ new \fbox{\textsc{map}} primitive)\\
  \code{(}\fbox{\textsc{map}}\code{ ($\lambda$ (z) (- z 1)))}\quad(program rewritten w/ new \fbox{\textsc{map}} primitive)\\
      \end{tabular}      
    };
    \draw [->](r1.south)--(m.north);
    \draw [->](r2.south)--(m.north);
    \node[fill=white] at ([yshift=0.5cm]m.north) {\textbf{Compress (MDL/Bayes objective)}};
    %% \visible<2-3>{
    %%   \node[fill=orange!50!white,align=center] at (r1) {\Huge $10^{14}$\\\Huge  refactorings };
    %%   \node[fill=orange!50!white,align=center] at (r2) {\Huge $10^{13}$\\\Huge  refactorings };
    %% }
    %% \visible<3>{
    %%   \node[fill=orange!50!white,align=center] at ([yshift=7cm]m.north) {\Huge version space: set of programs\\
    %%   \huge Lau 2003; Gulwani 2012};
    %%   }
    %% \visible<4>{
    %%   \node[fill=orange!50!white,align=center] at (r1) {\Huge $\leq 10^6$\\\Huge version spaces };
    %%   \node[fill=orange!50!white,align=center] at (r2) {\Huge $\leq 10^6$\\\Huge version spaces };
    %% }

@@

\subsection{Enumerative program search}\label{enumerationAppendix}

Our current implementation of \system takes the simple and generic strategy of enumerating programs in
descending order of their probability under either $(\mathcal{D},\theta)$ or $Q(p|x)$.
Algorithm~\ref{sampleProgram} and~\ref{recognitionSample}
specify procedures for sampling
from these distributions,
but not for enumerating from them.
We combine two different enumeration strategies,
which allowed us to build a massively parallel program enumerator:
\begin{itemize}
\item \textbf{Best-first search:} Best-first search maintains a heap of
  partial programs ordered by their
  probability --- here a partial program means a program whose syntax tree
  may contain unspecified `holes'. Best-first search
  is guaranteed to enumerate programs in decreasing order of their probability,
  and has memory requirements that in general grow exponentially as a function of the description length of programs in the heap (thus linearly as a function of run time).
\item \textbf{Depth-first search:} Depth first search
  recursively explores the space of
  execution traces through Algorithm~\ref{sampleProgram} and~\ref{recognitionSample},
  equivalently maintaining a stack of partial programs.  
  In general it does not enumerate programs
  in decreasing order of probability,
  but has memory requirements that grow linearly as a function of the description length of the programs in the stack (thus logarithmically as a function of run time).
\end{itemize}

Our parallel enumeration algorithm (Algorithm~\ref{enumerationAlgorithm})
first performs a best-first search
until the best-first heap
is much larger than the number of CPUs.
At this point,
it switches to performing many depth-first searches in parallel,
initializing a depth first search
with one of the
entries in the best-first heap.
Because depth-first search does not
produce programs in decreasing order of their probability,
we wrap this entire procedure up into an outer loop
that first enumerates programs whose description length is between
$0$ to $\Delta$,
then programs with description length between $\Delta$ and $2\Delta$,
then $2\Delta$ to $3\Delta$, etc.,
until a timeout is reached.
This is similar in spirit to iterative deepening depth first search~\cite{Russell:2003:AIM:773294}.
 \begin{algorithm}
   \caption{Parallel enumerative program search algorithm }
   \label{enumerationAlgorithm}
   \begin{algorithmic}[1]
     \State \textbf{function} enumerate$(\mu, T, \text{CPUs})$:
     \State {\bfseries Input:} Distribution over programs $\mu$, timeout $T$, CPU count
     \State \textbf{Output:} stream of programs in approximately descending order of probability under $\mu$
     \State \textbf{Hyperparameter:} nat increase rate $\Delta$\Comment{We set $\Delta = 1.5$}
     \State lowerBound$\gets 0$
     \While{total elapsed time $ < T$}
     \State heap$\gets$newMaxHeap()\Comment{Heap for best-first search}
     \State heap.insert$(\text{priority} = 0,\text{value} = \text{empty syntax tree})$\Comment{Initialize heap with start state of search space}
     \While{$0 < |\text{heap}|\leq 10\times \text{CPUs}$}\Comment{Each CPU will get approximately 10 jobs (a partial program)}
     \State priority, partialProgram $\gets$ heap.popMaximum()
     \If{partialProgram is finished}\Comment{Nothing more to fill in in the syntax tree}
     \If{$\text{lowerBound}\leq -\text{priority} < \text{lowerBound} + \Delta$}
     \State \textbf{yield }partialProgram
     \EndIf
     \Else
     \For{child$\in $children(partialProgram)}\Comment{children$(\cdot )$ fills in next random choice in syntax tree.}
     \If{$-\log \mu(\text{child}) < \text{lowerBound} + \Delta$}\Comment{Child's description length small enough}
     \State heap.insert$(\text{priority} = \log \mu(\text{child}),\text{value} = \text{child})$
     \EndIf
     \EndFor
     \EndIf
     \EndWhile
     \State \textbf{yield from }ParallelMap$_\text{CPUs}(\text{depthFirst}(\mu,T - \text{elapsed time}, \text{lowerBound}, \cdot ), \text{heap.values()})$%\Comment{Launch parallel workers}
     \State $\text{lowerBound}\gets\text{lowerBound} + \Delta$\Comment{Push up lower bound on MDL by $\Delta$}
     \EndWhile
     \Statex
     \State \textbf{function} depthFirst$(\mu,T,\text{lowerBound},\text{partialProgram})$: \Comment{Each worker does a depth first search. Enumerates completions of partialProgram whose MDL is between lowerBound and $\text{lowerBound} + \Delta$}
     \State stack$\gets$[partialProgram]
     \While{$\text{total elapsed time} < T$ and $\text{stack}$ is not empty}
     \State partialProgram$\gets$stack.pop()
     \If{partialProgram is finished}
     \If{$\text{lowerBound}\leq-\log \mu(\text{partialProgram}) < \text{lowerBound} + \Delta$}
     \State \textbf{yield }partialProgram
     \EndIf
     \Else
     \For{$\text{child}\in \text{children}(\text{partialProgram})$}
     \If{$-\log \mu(\text{child}) < \text{lowerBound} + \Delta$}\Comment{Child's description length small enough}
     \State stack.push$(\text{child})$
     \EndIf
     \EndFor
     \EndIf

     \EndWhile
   \end{algorithmic}
 \end{algorithm}
